---
title: "Class 11: Genome Informatics"
author: "Danika"
format: pdf
---

# Section 1: Proportion of G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39830100-39960101;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

> Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are
homozygous for the asthma associated SNP (G|G)?

14%

Now let's look at a different population. I picked the GBR.

> Q6. Back on the ENSEMBLE page, use the “search for a sample” field above to find the
particular sample HG00109. This is a male from the GBR population group. What is the
genotype for this sample?

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.)/ nrow(gbr) *100, 2)
```
30%

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Lets now dig into this further.

# Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.


> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

We can infer that the relative expression value for A/A is higher than that of the G/G and affects the expression of ORMDL3.

```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
ggplot(expr)+ aes(x=geno, y=exp, fill=geno)+
  geom_boxplot(notch=TRUE)
```















