---
title: "Class 17: Mini Project"
author: "Danika (A16020582)"
format: pdf
---
The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file 

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. What column details the total number of people fully vaccinated? 

vax$persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area?

vax$zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 

```{r}
vax$as_of_date[1]
```


```{r}
head(sort(vax$as_of_date, decreasing=FALSE))
```
2021-01-05

> Q4. What is the latest date in this dataset? 

```{r}
vax$as_of_date[nrow(vax)]
```

```{r}
head(sort(vax$as_of_date, decreasing=TRUE))
```
2023-02-28

We can use the skim() function fro a quick overview of a new data set like this

```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

13

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
table(is.na(vax$persons_fully_vaccinated))
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
n.missing <- sum(is.na(vax$persons_fully_vaccinated))
n.missing
```


```{r}
round((n.missing/nrow(vax))*100,2)
```

> Q8. [Optional]: Why might this data be missing?

Some zipcodes are not available (e.x. military) that are not recorded, so there are a few missing NAs

## Working with dates

The lubridate package makes working with dates and times in R muhc less of a pain. Let's have a first play with this package here.

```{r}
library(lubridate)
```

```{r}
today()
```
We can now magically do math with dates

```{r}
today() - ymd("2021-01-05")
```
How old am I

```{r}
today()-ymd("2001-02-06")
```
Let's treat the whole column as a date format

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```


> Q. How many days does the data set span.

```{r}
vax$as_of_date[nrow(vax)]-vax$as_of_date[1]
```


> Q9. How many days have passed since the last update of the dataset? 

```{r}
today()- vax$as_of_date[nrow(vax)]
```

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length((unique(vax$as_of_date)))
```
```{r}
library(dplyr)

n_distinct(vax$as_of_date)
```

## Working with Zip Codes

ZIP codes are also rather annoying things to work with as they are numeric but not in the conventional sense of doing math. 

Just like dates we have special packages to help us with ZIP codes.

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

```{r}
geocode_zip('92691')
```

```{r}
zip_distance('92037','92109')
```
More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:

```{r}
reverse_zipcode(c('92037', "92691") )
```


## Focus on the San Diego area

Let’s now focus in on the San Diego County area by restricting ourselves first to vax$county == "San Diego" entries. We have two main choices on how to do this. The first using base R the second using the dplyr package:

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego" , ]
nrow(sd)
```
It is time to revisit the most awesome **dyplr** package.

```{r}
library(dplyr)
sd <- filter(vax, county == "San Diego")

nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
nrow(sd.10)
```

Q. How many distinct zip codes are listed with population > 10,000

```{r}
n_distinct(sd.10$zip_code_tabulation_area)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
sd <- filter(vax, county == "San Diego")
                
n_distinct(sd$zip_code_tabulation_area)
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
ind <- which.max(sd$age12_plus_population)
sd$zip_code_tabulation_area[ind]
```

```{r}
reverse_zipcode("92154")
```

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”? 

```{r}
vax$as_of_date[nrow(vax)]
```

```{r}
##sd$as_of_date
sd.today <- filter(sd,as_of_date == "2023-02-28")
sd.today$percent_of_population_fully_vaccinated
```

```{r}
mean(sd.today$percent_of_population_fully_vaccinated, na.rm=T)
```



> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?

```{r}
hist(sd.today$percent_of_population_fully_vaccinated, breaks=20)
```

```{r}
library(ggplot2)

ggplot(sd.today) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=20) +
  labs(title="Vaccination rate across San Diego as of last week", 
       xlab= "Percent of Population Fully Vaxxed", 
       ylab="Count (ZIP code area)")
```

# Focus on UCSD/ La Jolla

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```


```{r}
ucplot <- ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title= "Vaccination Rate for La Jolla CA 92109",
  x="Date", y="Percent Vaccinated")
ucplot
```

## Comparing to similar sized areas

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2022-02-22”.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-11-15")

#head(vax.36)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
ave <- mean(vax.36$percent_of_population_fully_vaccinated)
ave
```

```{r}
ucplot + geom_hline(yintercept=ave,col="red", linetype=2)
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36)+
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(0,1)
```


> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
```{r}
x <- filter(vax.36, zip_code_tabulation_area %in% c ("92109", "92040"))
x$percent_of_population_fully_vaccinated
```


> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population>36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="red") +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccination",
       title="Vaccination Rate Across CA",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = 0.7213, linetype=2)
```
















