---
title: "Class 10: Halloween Candy Project"
author: "Danika"
format: pdf
---

##Importing Candy

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different candy types in this dataset.


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy[,2])
```
There are 38 fruity candy types.

## What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Twix", ]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
There is a useful `skim()` function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The "hist" and "skim_variable" variable/column looks to be on a different scale to the majority of the other columns in the dataset since it is not numeric.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

>Hint: look at the “Variable type” print out from the `skim()` function. Most variables (i.e. columns) are on the zero to one scale but not all. Some columns such as chocolate are exclusively either zero or one values.

A one represents if the candy contains chocolate. A zero represents the absense of chocolate in the candy.

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


> Q9. Is the distribution of winpercent values symmetrical?

The data is skewed to the right.

> Q10. Is the center of the distribution above or below 50%?

It is above 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
```
```{r}
mean(candy$winpercent[as.logical(candy$fruity)])
```
No, chocolate is higher ranked than fruit candy.


> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],candy$winpercent[as.logical(candy$fruity)])
```
The difference is statistically significant because the p value is less than 0.05.

## Overall Candy Ranking

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(-winpercent) %>% head(5)
```
> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent))+
  geom_col()
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```
> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starbursts

## Taking a look at price points

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head(candy[ord,c(11,12)], n=5)
```


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord,c(11,12)], n=5)
```


## Exploring Correlation Structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```


> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are the most anti correlated

> Q23. Similarly, what two variables are most positively correlated?

Bar and chocolate are the most positively correlated

## PCA

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


```{r}
library(plotly)
ggplotly(p)
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. Yes, this makes sense because fruity candies are usually hard and come in a bag or box of multiple.




